---
title: "Coursera Machine Learning Course Project: Writen Section"
output: html_document
---

##Executive Summary

The goal of this project is to build a predictive algorithm that will predict the exercise “classe variable” based on measurements from several sensors on the subject’s body and equipment. 

We are given a training set of 19622 observations and a test set of 20 observations. We break that up into a 60/40, training/validation set.  Variables with little variance or empty values are removed. Then random forest is used to create the algorithm.

Cross validation is performed using K-Fold method with 6 folds.

The algorithm is run against the 40% validation set to ascertain accuracy and out of sample error.

The model we created has 99.6% accuracy. And out of sample error rate of .4%

The final step is creating the prediction of exercise “classe variable” for the 20 test observations. 


##Loading data and requred library

```{r}
#load Librarys
library(randomForest)
library(randomForestSRC)
library(caret)
#Read Data
pmlTrain<- read.csv("pml-training.csv", sep=",")
pmlTest <- read.csv("pml-testing.csv", sep= ",") 
```

##Remove variables with low variability
```{r}
#set the seed to make your partition reproductible
set.seed(666)
#remove variables with near zero variance
nzv <- nearZeroVar(pmlTrain)
filtered <- pmlTrain[, -nzv]
filtered$X <- NULL
```

##Identify and remove variables with mostly NA values
```{r}
#View remaining variable
head(filtered)

#Remove varialbes that are mostly empty (value = NA)
inc.cols <- c("user_name", "num_window", "roll_belt", "pitch_belt", "yaw_belt", "gyros_belt_x", "gyros_belt_y", "gyros_belt_z", "accel_belt_x", "accel_belt_y", "accel_belt_z", "magnet_belt_x", "magnet_belt_y", "magnet_belt_z", "roll_arm", "pitch_arm", "yaw_arm", "total_accel_arm", "gyros_arm_x", "gyros_arm_y", "gyros_arm_z", "accel_arm_x", "accel_arm_y", "accel_arm_z", "magnet_arm_x", "magnet_arm_y", "magnet_arm_z", "roll_dumbbell", "pitch_dumbbell", "yaw_dumbbell", "gyros_dumbbell_x", "gyros_dumbbell_y", "gyros_dumbbell_z", "accel_dumbbell_x", "accel_dumbbell_y", "accel_dumbbell_z", "magnet_dumbbell_x", "magnet_dumbbell_y", "magnet_dumbbell_z", "roll_forearm", "pitch_forearm", "yaw_forearm", "total_accel_forearm", "gyros_forearm_x", "gyros_forearm_y", "gyros_forearm_z", "accel_forearm_x", "accel_forearm_y", "accel_forearm_z", "magnet_forearm_x", "magnet_forearm_y", "magnet_forearm_z", "classe")

filtered <- filtered[, inc.cols]
```

##Partition data into 60% train and 40% validation sets
```{r}
## Partition data into 60% train and 40% validation sets
smp_size <- floor(0.6 * nrow(pmlTrain))
train_ind <- sample(seq_len(nrow(pmlTrain)), size = smp_size)
filtered.train <- filtered[train_ind, ]
filtered.test <- filtered[-train_ind, ]
```

##Create model using Caret package with Random Forests method
```{r}
#Create model using random forests with caret
model <- train(filtered.train$classe ~.,data=filtered.train, method="rf")
save(model, file="model.rda")
load(file="model.rda")
```

##Perform cross validation using K-Folds method
```{r}
#Run crossvalidation using traning data
traincontrol <- trainControl(method = "cv", number = 6)
model_crossvalidation <- train(filtered.train$classe~.,data=filtered.train, method="rf",trControl=traincontrol)
save(model_crossvalidation, file="model_crossvalidation.rda")
load(file="model_crossvalidation.rda")
model_crossvalidation
```

##Create prediction using model and validation data set
```{r}
#test model using testing data
predTrainTest <- predict(model,filtered.test)
save(predTrainTest, file="predTrainTest.rda")
load(file="predTrainTest.rda")
table(predTrainTest,as.factor(filtered.test$classe))

confusionMatrix(predTrainTest,as.factor(filtered.test$classe))
```

##pmlTest set
Once the test set has been pre-processed a prediction is created using the model from previous steps. A function is this run to create the files required for submission.
```{r}
pmlTest.filtered <- pmlTest[, -nzv]
pmlTest.filtered$X <- NULL
inc.cols.test <- c("user_name", "num_window", "roll_belt", "pitch_belt", "yaw_belt", "gyros_belt_x", "gyros_belt_y", "gyros_belt_z", "accel_belt_x", "accel_belt_y", "accel_belt_z", "magnet_belt_x", "magnet_belt_y", "magnet_belt_z", "roll_arm", "pitch_arm", "yaw_arm", "total_accel_arm", "gyros_arm_x", "gyros_arm_y", "gyros_arm_z", "accel_arm_x", "accel_arm_y", "accel_arm_z", "magnet_arm_x", "magnet_arm_y", "magnet_arm_z", "roll_dumbbell", "pitch_dumbbell", "yaw_dumbbell", "gyros_dumbbell_x", "gyros_dumbbell_y", "gyros_dumbbell_z", "accel_dumbbell_x", "accel_dumbbell_y", "accel_dumbbell_z", "magnet_dumbbell_x", "magnet_dumbbell_y", "magnet_dumbbell_z", "roll_forearm", "pitch_forearm", "yaw_forearm", "total_accel_forearm", "gyros_forearm_x", "gyros_forearm_y", "gyros_forearm_z", "accel_forearm_x", "accel_forearm_y", "accel_forearm_z", "magnet_forearm_x", "magnet_forearm_y", "magnet_forearm_z")
pmlTest.filtered <- pmlTest.filtered[, inc.cols.test]

##Create Prediction for 20 observation test set
pmlTest.filtered.pred <- predict(model,pmlTest.filtered)

#Create files for prject Submission
 pml_write_files = function(x){
       n = length(x)
       for(i in 1:n){
             filename = paste0("problem_id_",i,".txt")
             write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
         }
   }
 pml_write_files(pmlTest.filtered.pred)
```
##Results
The accuracy of the model against the validation set is >99%. Crossvalidation also yelded high accuracy values.

The model predicted the exercise "classe" outcome for the test set with 100% accuracy
